---
title: 'DATA 607: Project 2'
author: "Keith Folsom"
date: "March 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Overview

DATA 607 Project 2 will take three wide datasets identified in the class discussion forum and use `tidyr` and `dplyr` as needed to tidy and transform the data into an analyzable format.

The datasets chosen for this project are:  
1. Credit Card Default Data -- UCI Machine Learning Repository  
2. New York City Leading Causes of Death -- NYC OpenData  
3. Total Population By Country -- World Bank  

####Setup

This analysis requires the following R packages:

1. dplyr
2. tidyr
3. stringr
4. RCurl
5. knitr
6. data.tree
7. ggplot2

```{r load packages, echo=FALSE}

suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(RCurl))
suppressPackageStartupMessages(library(knitr))
suppressWarnings(suppressMessages(library(data.tree)))
suppressPackageStartupMessages(library(ggplot2))

options(scipen=999)

```

##Dataset 1: Credit Card Default Data -- UCI Machine Learning Repository

####DataSet Information:

This research aimed at the case of customers default payments in Taiwan and was used to compare the predictive accuracy of probability of default among six data mining methods. 

The Credit Card Default Data on the UCI Machine Learning Repository can be found directly below:

http://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients

####Attribute Information:

This research employed a binary variable, default payment (Yes = 1, No = 0), as the response variable. 

This study reviewed the literature and used the following 23 variables as explanatory variables:    
**LIMIT_BAL**: Amount of the given credit (NT dollar): it includes both the individual consumer credit and his/her family (supplementary) credit.  
**SEX**      : Gender (1 = male; 2 = female).       
**EDUCATION**: Education (1 = graduate school; 2 = university; 3 = high school; 4 = others).   
**MARRIAGE** : Marital status (1 = married; 2 = single; 3 = others).   
**AGE**      : Age (year).   

_(X6 - X11)_  
**PAY_0 - PAY_6**: History of past payment. We tracked the past monthly payment records (from April to September, 2005) as follows: X6 = the repayment status in September, 2005; X7 = the repayment status in August, 2005; . . .;X11 = the repayment status in April, 2005. The measurement scale for the repayment status is: -1 = pay duly; 1 = payment delay for one month; 2 = payment delay for two months; . . .; 8 = payment delay for eight months; 9 = payment delay for nine months and above.   

_X12-X17_  
**BILL_AMT1 - BILL_AMT6**: Amount of bill statement (NT dollar). X12 = amount of bill statement in September, 2005; X13 = amount of bill statement in August, 2005; . . .; X17 = amount of bill statement in April, 2005.   

_X18-X23_  
**PAY_AMT1 - PAY_AMT6**: Amount of previous payment (NT dollar). X18 = amount paid in September, 2005; X19 = amount paid in August, 2005; . . .;X23 = amount paid in April, 2005.   



#####Load the Credit Card Default Data in CSV format from GitHub 

```{r, echo=FALSE}

URL <- getURL("https://raw.githubusercontent.com/kfolsom98/DATA607/master/Project%202/Data/default_of_credit_card_clients.csv")
credit_data_df <- read.csv(text = URL, header = TRUE, skip=1)

```

The dataset contains 30,000 observations and has 25 variables.

```{r, echo=FALSE}
str(credit_data_df)

```


####Start tidying the data

1. Fix what appears to be a typo in the field header PAY_0 

```{r}
names(credit_data_df)[names(credit_data_df) == "PAY_0"] <- "PAY_1"  

```

2.  Change codes to values for sex, education, and marriage.  Any observations associated with undocument code values will be removed.

```{r, echo = FALSE}

credit_data_df$SEX[credit_data_df$SEX == 1] <- "Male"
credit_data_df$SEX[credit_data_df$SEX == 2] <- "Female"

credit_data_df$EDUCATION[credit_data_df$EDUCATION == 1] <- "Graduate School"
credit_data_df$EDUCATION[credit_data_df$EDUCATION == 2] <- "University"
credit_data_df$EDUCATION[credit_data_df$EDUCATION == 3] <- "High School"

credit_data_df$EDUCATION <-  factor(credit_data_df$EDUCATION, 
                                levels = c("High School", 
                                           "University", 
                                           "Graduate School" ))

credit_data_df <- filter(credit_data_df, !EDUCATION == "4")
 

# some values for Education are not described in the mapping, remove these

credit_data_df <- filter(credit_data_df, !EDUCATION %in% c("0", "5", "6"))

credit_data_df$MARRIAGE[credit_data_df$MARRIAGE == 1] <- "Married"
credit_data_df$MARRIAGE[credit_data_df$MARRIAGE == 2] <- "Single"

credit_data_df <- filter(credit_data_df, !MARRIAGE %in% c("0", "3"))


credit_data_df <- rename(credit_data_df, DEFAULT_NEXT_PAYMENT_IND = default.payment.next.month)

credit_data_df_s <- credit_data_df %>% select( ID:AGE, PAY_1:PAY_AMT6, DEFAULT_NEXT_PAYMENT_IND) 

unique(credit_data_df$SEX)
unique(credit_data_df$EDUCATION)
unique(credit_data_df$MARRIAGE)

```

3.  Rename the columns in order to use `tidyr` to convert the dataset from wide to long format.

```{r, echo = FALSE}
cols <- colnames(credit_data_df_s)

cols[13:24] <- str_replace(cols[13:24], "_", "") %>%  str_replace("\\d+$", str_c("_", str_extract(cols[13:24], "\\d+$" )))

colnames(credit_data_df_s) <- cols

cols

```

4.  Convert from wide to long using  `tidyr` 

```{r}

tidy_credit_data_df <- credit_data_df_s %>%
                       gather(col, "VAL", -ID:-AGE, -DEFAULT_NEXT_PAYMENT_IND) %>%
                       separate(col, c("L1", "MTH"), sep="_") %>%
                       spread(L1, VAL) %>%
                       select(ID:MTH, BILLAMT, PAYAMT, PAY) 
```

**The resulting dataset looks like this:**

```{r, echo =FALSE}
kable(head(tidy_credit_data_df, 20))

```

5.  With the dataset in long format, create some derived fields:

Field Name |  Description 
-----------|--------------------------------------------------------------------------------
AMT_OWED   | Running or cumulative sum of bill amount - payment amount for each individual
AVG_6MTH_BAL| Mean value of AMT_OWED over a 6 month period
MISSED_PAYMENTS|Maxium number of missed payments recorded for the individual
BALANCE_TO_LIMIT_RATIO| Average 6 month balance divided by the individual's credit limit; note anything <= .3 is considered good
AGE_RANGE| Groups individuals into 10 year age grouping [20-29] through [70-79]

```{r, echo = FALSE}

#Rename some variables
tidy_credit_data_df <- tidy_credit_data_df %>% 
                       rename(CREDIT_LIMIT=LIMIT_BAL, MARITAL_STATUS=MARRIAGE) %>% 
                       arrange(ID, desc(MTH))

tidy_credit_data_df <-  tidy_credit_data_df %>% mutate(MADJ = BILLAMT - PAYAMT) 

tidy_credit_data_df <- tidy_credit_data_df %>% 
                       group_by(ID) %>% arrange(MTH)  %>%
                       mutate(AMT_OWED = cumsum(MADJ)) %>% 
                       select(-MADJ) %>% 
                       mutate(AVG_6MTH_BAL = round(mean(AMT_OWED), 3),
                              MISSED_PAYMENTS = max(PAY))

tidy_credit_data_df <- as.data.frame(tidy_credit_data_df)


credit_data_individual <- tidy_credit_data_df %>% 
                          filter(MTH==1) %>% 
                          select(ID:DEFAULT_NEXT_PAYMENT_IND, AVG_6MTH_BAL, MISSED_PAYMENTS) %>%
                          mutate(BALANCE_TO_LIMIT_RATIO = round(AVG_6MTH_BAL/CREDIT_LIMIT, 3))

credit_data_individual$AGE_RANGE[credit_data_individual$AGE %in% c(20:29)] <- "20-29" 
credit_data_individual$AGE_RANGE[credit_data_individual$AGE %in% c(30:39)] <- "30-39" 
credit_data_individual$AGE_RANGE[credit_data_individual$AGE %in% c(40:49)] <- "40-49" 
credit_data_individual$AGE_RANGE[credit_data_individual$AGE %in% c(50:59)] <- "50-59" 
credit_data_individual$AGE_RANGE[credit_data_individual$AGE %in% c(60:69)] <- "60-69" 
credit_data_individual$AGE_RANGE[credit_data_individual$AGE %in% c(70:79)] <- "70-79" 


credit_data_individual$AGE_RANGE <-  factor(credit_data_individual$AGE_RANGE, 
                                       levels = c("20-29", 
                                                  "30-39", 
                                                  "40-49", 
                                                  "50-59", 
                                                  "60-69",
                                                  "70-79"))
credit_data_individual$AGE <- NULL

```

**The resulting dataset looks like this:**

```{r, echo=FALSE}
kable(head(credit_data_individual, 10))

```


6.  Using the dataset just created and stored in `credit_data_individual`, create an aggregate dataset for the different group combinations of Sex, Age Range, Marital Status, and Education.

Aggregation Level |  Description 
-----------|--------------------------------------------------------------------------------
LVL1  | Top Level - Aggregates to Age Range  
LVL2  | Aggregates to Age Range and Sex
LVL3  | Aggregates to Age Range, Sex, Marital Satus 
LVL4  | Bottom Level - Aggregates to Age Range, Sex, Marital Satus, Education 

```{r, echo=FALSE}
credit_data_aggr <- 
                   credit_data_individual %>% 
                   select(-ID) %>% 
                   group_by(SEX, MARITAL_STATUS, EDUCATION, AGE_RANGE) %>% 
                   summarise(GROUP_TOTAL = n(),
                           GOOD_STANDING_TOTAL = sum(ifelse(BALANCE_TO_LIMIT_RATIO <= .3, 1, 0)),
                           AVG_CREDIT_LIMIT = round(mean(CREDIT_LIMIT), 0),
                           TOTAL_DEFAULT = sum(DEFAULT_NEXT_PAYMENT_IND),
                           CNT_0_MISSED_PAYMENTS   = sum(ifelse(MISSED_PAYMENTS <= 0, 1, 0)),
                           CNT_MISSED_PAYMENTS = sum(ifelse(MISSED_PAYMENTS %in% c(1:9), 1, 0))) %>%
                    mutate(PCT_IN_GROUP_GOOD_CREDIT = round(GOOD_STANDING_TOTAL/GROUP_TOTAL, 2),
                           PCT_IN_GROUP_DEFAULT_RATE = round(TOTAL_DEFAULT/ GROUP_TOTAL, 2), 
                           PCT_IN_GROUP_0_MISSED_PAYMENTS = round(CNT_0_MISSED_PAYMENTS/GROUP_TOTAL, 2 ),
                           PCT_IN_GROUP_MISSED_PAYMENT = round(CNT_MISSED_PAYMENTS/GROUP_TOTAL, 2 ))

credit_data_aggr <- filter(credit_data_aggr, GROUP_TOTAL > 10) %>%
                    select(-GOOD_STANDING_TOTAL, -TOTAL_DEFAULT, -CNT_0_MISSED_PAYMENTS, -CNT_MISSED_PAYMENTS)
       
#format to currency
credit_data_aggr$FORMATTED_AVG_CREDIT_LIMIT <- paste("$",format(credit_data_aggr$AVG_CREDIT_LIMIT, big.mark=","), sep="")
credit_data_aggr <- as.data.frame(credit_data_aggr)

credit_data_aggr_lvl4 <- credit_data_aggr

# ----------------------------------------------------------
# Level 3 - Aggregate to Sex, Age Range, and Marital Status
# ----------------------------------------------------------
credit_data_aggr_lvl3 <- 
                   credit_data_individual %>% 
                   select(-ID) %>% 
                   group_by(SEX, AGE_RANGE, MARITAL_STATUS ) %>% 
                   summarise(GROUP_TOTAL = n(),
                           GOOD_STANDING_TOTAL = sum(ifelse(BALANCE_TO_LIMIT_RATIO <= .3, 1, 0)),
                           AVG_CREDIT_LIMIT = round(mean(CREDIT_LIMIT), 0),
                           TOTAL_DEFAULT = sum(DEFAULT_NEXT_PAYMENT_IND),
                           CNT_0_MISSED_PAYMENTS   = sum(ifelse(MISSED_PAYMENTS <= 0, 1, 0)),
                           CNT_MISSED_PAYMENTS = sum(ifelse(MISSED_PAYMENTS %in% c(1:9), 1, 0))) %>%
                    mutate(PCT_IN_GROUP_GOOD_CREDIT = round(GOOD_STANDING_TOTAL/GROUP_TOTAL, 2),
                           PCT_IN_GROUP_DEFAULT_RATE = round(TOTAL_DEFAULT/ GROUP_TOTAL, 2), 
                           PCT_IN_GROUP_0_MISSED_PAYMENTS = round(CNT_0_MISSED_PAYMENTS/GROUP_TOTAL, 2 ),
                           PCT_IN_GROUP_MISSED_PAYMENT = round(CNT_MISSED_PAYMENTS/GROUP_TOTAL, 2 ))

credit_data_aggr_lvl3 <- filter(credit_data_aggr_lvl3, GROUP_TOTAL > 10) %>%
                        select(-GOOD_STANDING_TOTAL, -TOTAL_DEFAULT, -CNT_0_MISSED_PAYMENTS, -CNT_MISSED_PAYMENTS)

#format to currency
credit_data_aggr_lvl3$FORMATTED_AVG_CREDIT_LIMIT <- paste("$",format(credit_data_aggr_lvl3$AVG_CREDIT_LIMIT, big.mark=","), sep="")
credit_data_aggr_lvl3 <- as.data.frame(credit_data_aggr_lvl3)

# create placeholder column for education
credit_data_aggr_lvl3 <- credit_data_aggr_lvl3 %>%
                         mutate(EDUCATION = NA)  %>% 
                         select(SEX, MARITAL_STATUS, EDUCATION, AGE_RANGE, GROUP_TOTAL:FORMATTED_AVG_CREDIT_LIMIT) 

credit_data_aggr_lvl3$EDUCATION <- as.factor(credit_data_aggr_lvl3$EDUCATION)

# ----------------------------------------------------------
# Level 2 - Aggregate to Sex and Age Range 
# ----------------------------------------------------------
credit_data_aggr_lvl2 <- 
                   credit_data_individual %>% 
                   select(-ID) %>% 
                   group_by(AGE_RANGE, SEX) %>% 
                   summarise(GROUP_TOTAL = n(),
                           GOOD_STANDING_TOTAL = sum(ifelse(BALANCE_TO_LIMIT_RATIO <= .3, 1, 0)),
                           AVG_CREDIT_LIMIT = round(mean(CREDIT_LIMIT), 0),
                           TOTAL_DEFAULT = sum(DEFAULT_NEXT_PAYMENT_IND),
                           CNT_0_MISSED_PAYMENTS   = sum(ifelse(MISSED_PAYMENTS <= 0, 1, 0)),
                           CNT_MISSED_PAYMENTS = sum(ifelse(MISSED_PAYMENTS %in% c(1:9), 1, 0))) %>%
                    mutate(PCT_IN_GROUP_GOOD_CREDIT = round(GOOD_STANDING_TOTAL/GROUP_TOTAL, 2),
                           PCT_IN_GROUP_DEFAULT_RATE = round(TOTAL_DEFAULT/ GROUP_TOTAL, 2), 
                           PCT_IN_GROUP_0_MISSED_PAYMENTS = round(CNT_0_MISSED_PAYMENTS/GROUP_TOTAL, 2 ),
                           PCT_IN_GROUP_MISSED_PAYMENT = round(CNT_MISSED_PAYMENTS/GROUP_TOTAL, 2 ))

credit_data_aggr_lvl2 <- filter(credit_data_aggr_lvl2, GROUP_TOTAL > 10) %>%
                        select(-GOOD_STANDING_TOTAL, -TOTAL_DEFAULT, -CNT_0_MISSED_PAYMENTS, -CNT_MISSED_PAYMENTS)


#format to currency
credit_data_aggr_lvl2$FORMATTED_AVG_CREDIT_LIMIT <- paste("$",format(credit_data_aggr_lvl2$AVG_CREDIT_LIMIT, big.mark=","), sep="")
credit_data_aggr_lvl2 <- as.data.frame(credit_data_aggr_lvl2)

# create placeholder column for education and marital status
credit_data_aggr_lvl2 <- credit_data_aggr_lvl2 %>%
                         mutate(EDUCATION = NA, MARITAL_STATUS = NA)  %>% 
                         select(SEX, MARITAL_STATUS, EDUCATION, AGE_RANGE, GROUP_TOTAL:FORMATTED_AVG_CREDIT_LIMIT) 

credit_data_aggr_lvl2$EDUCATION <- as.factor(credit_data_aggr_lvl2$EDUCATION)
credit_data_aggr_lvl2$MARITAL_STATUS <- as.character.Date(credit_data_aggr_lvl2$MARITAL_STATUS)

# ----------------------------------------------------------
# Level 2 - Aggregate Age Range 
# ----------------------------------------------------------
credit_data_aggr_lvl1 <- 
                   credit_data_individual %>% 
                   select(-ID) %>% 
                   group_by(AGE_RANGE) %>% 
                   summarise(GROUP_TOTAL = n(),
                           GOOD_STANDING_TOTAL = sum(ifelse(BALANCE_TO_LIMIT_RATIO <= .3, 1, 0)),
                           AVG_CREDIT_LIMIT = round(mean(CREDIT_LIMIT), 0),
                           TOTAL_DEFAULT = sum(DEFAULT_NEXT_PAYMENT_IND),
                           CNT_0_MISSED_PAYMENTS   = sum(ifelse(MISSED_PAYMENTS <= 0, 1, 0)),
                           CNT_MISSED_PAYMENTS = sum(ifelse(MISSED_PAYMENTS %in% c(1:9), 1, 0))) %>%
                    mutate(PCT_IN_GROUP_GOOD_CREDIT = round(GOOD_STANDING_TOTAL/GROUP_TOTAL, 2),
                           PCT_IN_GROUP_DEFAULT_RATE = round(TOTAL_DEFAULT/ GROUP_TOTAL, 2), 
                           PCT_IN_GROUP_0_MISSED_PAYMENTS = round(CNT_0_MISSED_PAYMENTS/GROUP_TOTAL, 2 ),
                           PCT_IN_GROUP_MISSED_PAYMENT = round(CNT_MISSED_PAYMENTS/GROUP_TOTAL, 2 ))


credit_data_aggr_lvl1 <- filter(credit_data_aggr_lvl1, GROUP_TOTAL > 10) %>%
                        select(-GOOD_STANDING_TOTAL, -TOTAL_DEFAULT, -CNT_0_MISSED_PAYMENTS, -CNT_MISSED_PAYMENTS)

#format to currency
credit_data_aggr_lvl1$FORMATTED_AVG_CREDIT_LIMIT <- paste("$",format(credit_data_aggr_lvl1$AVG_CREDIT_LIMIT, big.mark=","), sep="")
credit_data_aggr_lvl1 <- as.data.frame(credit_data_aggr_lvl1)

# create placeholder column for education and marital status
credit_data_aggr_lvl1 <- credit_data_aggr_lvl1 %>%
                         mutate(EDUCATION = NA, MARITAL_STATUS = NA, SEX = NA)  %>% 
                         select(SEX, MARITAL_STATUS, EDUCATION, AGE_RANGE, GROUP_TOTAL:FORMATTED_AVG_CREDIT_LIMIT) 

credit_data_aggr_lvl1$EDUCATION <- as.factor(credit_data_aggr_lvl1$EDUCATION)
credit_data_aggr_lvl1$MARITAL_STATUS <- as.character.Date(credit_data_aggr_lvl1$MARITAL_STATUS)
credit_data_aggr_lvl1$SEX <- as.character.Date(credit_data_aggr_lvl1$SEX)

```

###Visualize the groups using the  `data.tree` package:

**Field Definition:**

Aggregation Level |  Description 
-----------|--------------------------------------------------------------------------------
n  | Count of individuals in the groups 
Avg Limit  | Average Credit Limit of the group
% Good Credit  | Percentage representation of how many individuals in the group have a <= .3 limit-to-balance ratio
% Default | Percentage representation of how many individuals in the group are predicted to default

**Level 1: Summarized to Age Range**

``` {r, echo=FALSE}

credit_data_tree <- select(credit_data_aggr_lvl1, AGE_RANGE, GROUP_TOTAL, FORMATTED_AVG_CREDIT_LIMIT, 
                           PCT_IN_GROUP_GOOD_CREDIT, PCT_IN_GROUP_DEFAULT_RATE)

credit_data_tree <- 
                  credit_data_tree %>% 
                  rename("n" = GROUP_TOTAL) %>%
                  rename("Avg Limit"   = FORMATTED_AVG_CREDIT_LIMIT) %>%
                  rename("% Good Credit" = PCT_IN_GROUP_GOOD_CREDIT)  %>%
                  rename("% Default" = PCT_IN_GROUP_DEFAULT_RATE)


credit_data_tree$pathString <- paste("All", 
                            credit_data_tree$AGE_RANGE, 
                            sep = "/")
                            
groups <- data.tree::as.Node(credit_data_tree)

print(groups, "n", "Avg Limit", "% Good Credit", "% Default")

```

**Level 2: Summarized to Age Range and Sex **

``` {r, echo=FALSE} 
credit_data_tree <- select(credit_data_aggr_lvl2, AGE_RANGE, SEX, 
                           GROUP_TOTAL, FORMATTED_AVG_CREDIT_LIMIT, 
                           PCT_IN_GROUP_GOOD_CREDIT, PCT_IN_GROUP_DEFAULT_RATE)

credit_data_tree <- 
                  credit_data_tree %>% 
                  rename("n" = GROUP_TOTAL) %>%
                  rename("Avg Limit"   = FORMATTED_AVG_CREDIT_LIMIT) %>%
                  rename("% Good Credit" = PCT_IN_GROUP_GOOD_CREDIT)  %>%
                  rename("% Default" = PCT_IN_GROUP_DEFAULT_RATE)


credit_data_tree$pathString <- paste("All", 
                            credit_data_tree$AGE_RANGE, 
                            credit_data_tree$SEX, 
                            sep = "/")
                            
groups <- data.tree::as.Node(credit_data_tree)
print(groups, "n", "Avg Limit", "% Good Credit", "% Default")

```

**Level 3: Summarized to Age Range, Sex, and Marital Status **

``` {r, echo=FALSE} 

credit_data_tree <- select(credit_data_aggr_lvl3, AGE_RANGE, SEX, MARITAL_STATUS, GROUP_TOTAL, FORMATTED_AVG_CREDIT_LIMIT, PCT_IN_GROUP_GOOD_CREDIT, PCT_IN_GROUP_DEFAULT_RATE)

credit_data_tree <- 
                  credit_data_tree %>% 
                  rename("n" = GROUP_TOTAL) %>%
                  rename("Avg Limit"   = FORMATTED_AVG_CREDIT_LIMIT) %>%
                  rename("% Good Credit" = PCT_IN_GROUP_GOOD_CREDIT)  %>%
                  rename("% Default" = PCT_IN_GROUP_DEFAULT_RATE)


credit_data_tree$pathString <- paste("All", 
                            credit_data_tree$AGE_RANGE, 
                            credit_data_tree$SEX, 
                            credit_data_tree$MARITAL_STATUS,
                            sep = "/")
                            
groups <- data.tree::as.Node(credit_data_tree)
print(groups, "n", "Avg Limit", "% Good Credit", "% Default")
```


**Level 4: Summarized to Age Range, Sex, Marital Status, and Education **

``` {r, echo=FALSE} 
credit_data_tree <- select(credit_data_aggr_lvl4, AGE_RANGE, SEX, MARITAL_STATUS, EDUCATION, GROUP_TOTAL, FORMATTED_AVG_CREDIT_LIMIT, PCT_IN_GROUP_GOOD_CREDIT, PCT_IN_GROUP_DEFAULT_RATE)

credit_data_tree <- 
                  credit_data_tree %>% 
                  rename("n" = GROUP_TOTAL) %>%
                  rename("Avg Limit"   = FORMATTED_AVG_CREDIT_LIMIT) %>%
                  rename("% Good Credit" = PCT_IN_GROUP_GOOD_CREDIT)  %>%
                  rename("% Default" = PCT_IN_GROUP_DEFAULT_RATE)


credit_data_tree$pathString <- paste("All", 
                            credit_data_tree$AGE_RANGE, 
                            credit_data_tree$SEX, 
                            credit_data_tree$MARITAL_STATUS,
                            credit_data_tree$EDUCATION,
                            sep = "/")
                            
groups <- data.tree::as.Node(credit_data_tree)

print(groups, "n", "Avg Limit", "% Good Credit", "% Default")

```



###Analysis 

**Which group has the highest average credit limit?**

```{r, echo = FALSE}

kable(
credit_data_aggr %>% select(SEX:AGE_RANGE, FORMATTED_AVG_CREDIT_LIMIT, AVG_CREDIT_LIMIT) %>%
                    arrange(desc(AVG_CREDIT_LIMIT)) %>% 
                    filter(row_number() == 1 ) %>% 
                    select(-AVG_CREDIT_LIMIT))

```

**Which group has the lowest average credit limit?**

```{r, echo = FALSE}

kable(
      credit_data_aggr %>% 
       select(SEX:AGE_RANGE, FORMATTED_AVG_CREDIT_LIMIT, AVG_CREDIT_LIMIT) %>% 
       arrange(AVG_CREDIT_LIMIT) %>% 
        filter(row_number() == 1) %>% select(-AVG_CREDIT_LIMIT))

```

**Which group is comprised of highest percentage of people who have a balance-to-limit rating less than or equal to 30%?**

```{r, echo = FALSE}

kable(
credit_data_aggr %>% 
    select(SEX:AGE_RANGE, PCT_IN_GROUP_GOOD_CREDIT) %>% 
    arrange(desc(PCT_IN_GROUP_GOOD_CREDIT)) %>%  
    filter(row_number() == 1))

```

**Which group has the lowest utilization or balance-to-limit rating?**

```{r, echo = FALSE}

kable(
credit_data_aggr %>% 
    select(SEX:AGE_RANGE, PCT_IN_GROUP_GOOD_CREDIT) %>% 
    arrange(PCT_IN_GROUP_GOOD_CREDIT) %>%  
    filter(row_number() == 1))

```

**Which group is the most likely to predicted to default?**

```{r, echo = FALSE}
kable(
credit_data_aggr %>% select(SEX:AGE_RANGE, PCT_IN_GROUP_DEFAULT_RATE) %>% 
        arrange(desc(PCT_IN_GROUP_DEFAULT_RATE)) %>% 
        filter(row_number() == 1))


```

**Which group has the highest amount of debt, is the most likely to default, and is the most likely to miss a payment?**

```{r, echo = FALSE}

d1 <- credit_data_aggr %>% 
      select(SEX:AGE_RANGE, PCT_IN_GROUP_GOOD_CREDIT) %>% 
      arrange(PCT_IN_GROUP_GOOD_CREDIT) %>% 
      filter(row_number() < 6) %>% 
      select(SEX:AGE_RANGE)

d2 <- credit_data_aggr %>% 
      select(SEX:AGE_RANGE, PCT_IN_GROUP_DEFAULT_RATE) %>% 
      arrange(desc(PCT_IN_GROUP_DEFAULT_RATE)) %>% 
      filter(row_number() < 6) %>% 
      select(SEX:AGE_RANGE)

d3 <- credit_data_aggr %>% 
      select(SEX:AGE_RANGE, PCT_IN_GROUP_MISSED_PAYMENT) %>% 
      arrange(desc(PCT_IN_GROUP_MISSED_PAYMENT)) %>% 
      filter(row_number() < 6) %>% 
      select(SEX:AGE_RANGE)

kable(intersect(d1, d2, d3))

```

**Which group has the lowest amount of debt, is the least predicted to default, and is not likely to miss a payment?**


```{r, echo = FALSE}

d1 <- credit_data_aggr %>% 
      select(SEX:AGE_RANGE, PCT_IN_GROUP_GOOD_CREDIT) %>% 
      arrange(desc(PCT_IN_GROUP_GOOD_CREDIT)) %>% 
      filter(row_number() < 6) %>% 
      select(SEX:AGE_RANGE)

d2 <- credit_data_aggr %>% 
      select(SEX:AGE_RANGE, PCT_IN_GROUP_DEFAULT_RATE) %>% 
      arrange(PCT_IN_GROUP_DEFAULT_RATE) %>% 
      filter(row_number() < 6) %>% 
      select(SEX:AGE_RANGE)

d3 <- credit_data_aggr %>% 
      select(SEX:AGE_RANGE, PCT_IN_GROUP_MISSED_PAYMENT) %>% 
      arrange(PCT_IN_GROUP_MISSED_PAYMENT) %>% 
      filter(row_number() < 6) %>% 
      select(SEX:AGE_RANGE)

kable(intersect(d1, d2, d3))

```



##Dataset 2: New York City Leading Causes of Death -- NYC OpenData


####DataSet Information:

This dataset collects the leading causes of death for New York City by ethnicity and sex from 2007 - 2011. 

The New York City Leading Causes of Death dataset can be found directly below:

https://data.cityofnewyork.us/Health/New-York-City-Leading-Causes-of-Death/jb7j-dtam

####Attribute Information:

The following attributes are captured in the dataset:

**Year**   
**Ethnicity**    
**Sex**        
**Cause of Death**   
**Count**   
**Percent**     

####Load the NYC Leading Causes of Death datset in CSV format from GitHub 

```{r, echo = FALSE}

URL <- getURL("https://raw.githubusercontent.com/kfolsom98/DATA607/master/Project%202/Data/New_York_City_Leading_Causes_of_Death.csv")
nyc_deaths_df <- read.csv(text = URL, header = TRUE, stringsAsFactors = FALSE)

```

Looking at the dataset, there are 2880 observations with 6 variables.

```{r, echo = FALSE}

nyc_deaths_df <- as.data.frame(nyc_deaths_df)

str(nyc_deaths_df)

```

However, there appears to be duplicate entries in the data.  

```{r, echo = FALSE}
kable( 
       head (
               nyc_deaths_df %>% 
                  group_by(Year, Ethnicity, Sex, Cause.of.Death, Count, Percent) %>% 
                  mutate(rn = row_number()) %>% filter(rn > 1) %>%
                  arrange(Year, Ethnicity, Sex, Cause.of.Death, Count, Percent), 20))
    
    
```

Remove the duplicate entries:

```{r }

nyc_deaths_df <-  nyc_deaths_df %>% 
                  group_by(Year, Ethnicity, Sex, Cause.of.Death, Count, Percent) %>% 
                  filter(row_number() == 1)

dim(nyc_deaths_df)


```

Removing the duplicate reduces the data down to **960** observations.


###Analysis:

**Show the leading causes of death each year for men and women.**

```{r}

kable(nyc_deaths_df %>% 
           select(Year, Sex, Cause.of.Death, Count) %>% 
           group_by(Year, Sex) %>% 
           select(Year, Sex, Cause.of.Death, Count) %>% arrange(Year, Sex) %>%
           top_n(1) )

```

**Show the leading causes of death each year for each ethnic group.**

```{r}

kable(nyc_deaths_df %>% 
         select(Year, Ethnicity, Cause.of.Death, Count) %>% 
         group_by(Year, Ethnicity) %>%
         select(Year, Ethnicity, Cause.of.Death, Count) %>% arrange(Year, Ethnicity) %>%
         top_n(1))

```


**Calculate which cause of death has declined the most and which has increased the most in the years given.**

```{r}

wide <- nyc_deaths_df %>% 
        select(Year, Cause.of.Death, Count) %>% 
        group_by(Year, Cause.of.Death) %>%
        summarise(Total = sum(Count)) %>% 
        spread(Year, Total)  %>% 
        mutate(Pct.Change = round((`2011`-`2007`)/`2007`, 2) *100)

wide <- wide[complete.cases(wide), ]

kable(wide %>% 
           arrange(Pct.Change) %>% 
           filter(row_number() == 1))
               
kable(wide %>% 
          arrange(desc(Pct.Change)) %>% 
          filter(row_number() == 1))

```

**Calculate which cause of death has remained stable over the years given.**

The answer to this question uses the percentage change when comparing year 2007 to year 2011.  There some fluctuations up and down in the amount of change in the years between.


```{r}

kable(wide %>% filter(Pct.Change == 0))

```


##Dataset 3: Total Population By Country

####DataSet Information:

Total population is based on the de facto definition of population, which counts all residents regardless of legal status or citizenship--except for refugees not permanently settled in the country of asylum, who are generally considered part of the population of their country of origin.

The population data is captured for each country from 1960 - 2014.

The World Bank Total Populdation dataset can be found directly below:

http://data.worldbank.org/indicator/SP.POP.TOTL

####Attribute Information:

The following attributes are captured in the dataset:

**Country Name:** Name of the Country
**Country Code:** Country's 3-digit ISO code	
**Indicator Name:** Constant value - "Population, total"
**Indicator Code:** Constant value - SP.POP.TOTL
**Year [1960 - 2015]:** Population of the country for the given year; 2015 does not contain any values for population
   
 
#####Load the World Bank Total Population Data in CSV format from GitHub 

```{r, echo=FALSE}

URL <- getURL("https://raw.githubusercontent.com/kfolsom98/DATA607/master/Project%202/Data/Total_Population.csv")
population_df <- read.csv(text = URL, header = TRUE, skip=4, stringsAsFactors = FALSE)

```

The dataset contains 248 observations and has 61 variables.

```{r, echo=FALSE}
str(population_df)

```

Remove variables that will not be used or are not valued -- Indicator Name, Indicator Code, X2015, and X (which appears after processing in the CSV file).

```{r}
population_df$Indicator.Name <- NULL
population_df$Indicator.Code <- NULL
population_df$X  <- NULL
population_df$X2015  <- NULL
```
  
Remove incomplete cases

```{r}
population_df <- population_df[complete.cases(population_df), ] 
  
``` 

Convert the Total Population dataset from wide to long format:

```{r}

tidy_population_df <- population_df %>%
                      gather(col, "Population", -Country.Code, -Country.Name) %>%
                      separate(col, c("X", "Year"), sep="X") %>%
                      select(-X) %>% 
                      arrange(Country.Code, Year)

tidy_population_df$Year <- as.numeric(tidy_population_df$Year)

tidy_population_df <- as.data.frame(tidy_population_df)
```

The resulting dataframe now has one row per year and population value for a country.  In total, there are 13,640 observations with just four variables.

```{r, echo = FALSE}

kable(filter(tidy_population_df, row_number() < 20))

```


###Analysis:

Determine which countries have had the most significant change in terms of straight percent population change, comparing 1960 to 2014.

```{r} 
population_aggr <- tidy_population_df %>%
                   na.omit() %>% 
                   group_by(Country.Code) %>% 
                   mutate(min_year = min(Year),
                   max_year = max(Year)) %>%
                   filter(Year == min_year | Year == max_year) %>%  
                   mutate(rn = str_c("M", row_number())) %>%
                   select(Country.Name, Country.Code, Population, rn) %>% 
                   spread(rn, Population) %>% 
                   rename(Min.Year=M1, Max.Year=M2) %>%
                   mutate(Pct.Change = round(((Max.Year - Min.Year)/Min.Year) * 100, 2)) 


population_aggr <- as.data.frame(population_aggr)

```

Using this resultset, determine which countries have experiences the most total growth and which experienced the least amount of growth:

```{r} 

countries <- filter(population_aggr, Pct.Change==min(Pct.Change) |Pct.Change==max(Pct.Change))

kable(countries)

``` 

We see that Bulgaria has seen its population drop the most significantly over the years, while United Arab Emirates has seen a population explosion.


To drill into these two countries more, let's look at the percent population change by decade.

```{r, echo = FALSE} 

population_by_decade <- 
                       tidy_population_df %>% 
                       filter(Year <= 2009) %>% 
                       mutate(Decade = str_c(str_sub(Year, 1, 3), 0)) %>%
                       group_by(Country.Code, Decade) %>% 
                       filter(Year == min(Year) | Year == max(Year)) %>%  
                       mutate(rn = str_c("M", row_number())) %>%
                       select(Country.Name, Country.Code, Population, Decade, rn) %>% 
                       spread( rn, Population) %>% 
                       rename(Min.Year=M1, Max.Year=M2) %>%
                       mutate(Pct.Change = round(((Max.Year - Min.Year)/Min.Year) * 100, 2)) %>% 
                       select(-Min.Year, -Max.Year) %>%
                       spread( Decade, Pct.Change) %>%
                       rename(Pct.Change.1960s = `1960`, 
                              Pct.Change.1970s = `1970`, 
                              Pct.Change.1980s = `1980`, 
                              Pct.Change.1990s = `1990`, 
                              Pct.Change.2000s = `2000`)

population_by_decade <- as.data.frame(population_by_decade)

```

The resulting dataset looks like:

```{r, echo = FALSE}

kable(head(population_by_decade, 10))

```

Looking at the change for Bulgaria and United Arab Emirates specifically:

```{r, echo = FALSE}
kable(filter(population_by_decade, Country.Code %in% c("BGR", "ARE")))

```

Plotting the population by year for both countries:

```{r, echo = FALSE}
line <- tidy_population_df %>% filter(Country.Code  %in% c("BGR", "ARE") ,
             Year %in% c(1960, 1965, 1970, 1975, 1980, 1985, 1990, 1995, 2000, 2005, 2010, 2014) )

ggplot(data=line, aes(x=factor(Year), y=Population, group=Country.Code, colour=Country.Name)) +
    geom_line() +
    geom_point() +
    ggtitle("Population By Year - Bulgaria vs. United Arab Emirates") +
    xlab("Year") + ylab("Total Population") 

```

For relative growth by decade:

```{r, echo = FALSE}
bar <- population_by_decade %>% 
       filter(Country.Code  %in% c("BGR", "ARE")) %>%
       gather(col, "Pct.Growth", -Country.Name, -Country.Code) %>%
       separate(col, c("X", "Decade"), sep="Pct.Change.")   %>%
        mutate(Decade = str_extract(Decade, "[0-9]+")) %>% select(-X)

# Use the original data frame, but put factor() directly in the plot specification
ggplot(data=bar, aes(x=factor(Decade), y=Pct.Growth, fill=Country.Name)) +
    geom_bar(stat="identity", position=position_dodge()) +
    ggtitle("Percentage Comparison - Bulgaria vs. United Arab Emirates") +
    xlab("Decade") + ylab("Percent Growth") 
       
```
  
  
  


