---
title: 'DATA607: Week 4 Assignment - R Character Manipulation and Date Processing'
author: "Keith Folsom"
date: "February 18, 2016"
output: html_document
---

The assignment for Week 4 will focus on regular expressions and requires the stringr package to be loaded.

```{r}
library(stringr)

```

###3. Copy the introductory example.  The vector `name` stores the extracted columns.

```{r, eval = TRUE}
# create the name vector with the Simpsons characters' names 

name <- c("Moe Szyslak", "Burns, C. Montgomery", "Rev. Timothy Lovejoy", "Ned Flanders", "Simpson, Homer", "Dr. Julius Hibbert")

# view the names
name

```
####(a) Use the tools of this chapter to rearrange the vector so that all elements conform to the standard first_name last_name.

To do this, I created a function to loop through the elements in the vector to handle each different scenario in the name format.

```{r, eval = TRUE}

#################################################################################
# Function   : name_standardize
# Description: When passed a name vector with varying name formats
#              this function will return a string in the format of:
#                  Title First_Name Middle_Name Last_Name
#              The parameter remove_title is defaulted to TRUE; set to FALSE to
#              leave the title in the name vector returned.
#################################################################################
name_standardize <- function(names_in, remove_title = TRUE) {
    
        return_names <- names_in 
        
        # remove title or prefix from name string before processing
        # This currently include Rev or Dr but could be expanded to include Mr./Mrs.
        if (remove_title) {
            
            return_names <- str_replace(return_names, "Rev. |Dr. ", "")
        }
        
        for (i in 1:length(return_names)) {
            
              # if the first word is followed by a comma, assume this is the last name
              # and adjust to the end of the string
            
              if (str_detect(return_names[i], ",") == TRUE ) {
                  
                  first_name <- str_extract(return_names[i], "(?<=, ).*?(?=$)")
                  last_name  <- str_extract(return_names[i], "(?<=\\w?).*?(?=,)")
                  
                  return_names[i] <- str_c(first_name, " ", last_name)
              }
            
        }    
        
        return(return_names)
}

#################################################################################
# Function   : remove_middle_name
# Description: When passed a name string in the format of:
#                     first_name middle_name last_name
#              removes the middle name and returns first_name last_name
#################################################################################
remove_middle_name <- function(names_in) {
    
       return_names <- names_in  # initialize the return vector 
    
        # handle any names that include a middle name
        # this block of code removes the middle name from the string
        for (i in 1:length(return_names)) {
            
            if (length(unlist(str_split(return_names[i], " "))) == 3 ) {
                  
                 first_name <- str_extract(return_names[i], "^(\\w+)")
                 last_name  <- str_extract(return_names[i], "(\\w+)$")
                 return_names[i] <- str_c(first_name, " ", last_name)
                  
             }
        }
        
        
        return(return_names)
}

```

Applying the function `name_standardize` to the vector of names will return the names in the more standard format of:  

_Title First_Name Middle_Name Last_Name_   

*Note. the parameter `remove_title` is defaulted to TRUE; change this paramater to FALSE will leave the title in place

```{r, eval = TRUE}

standard_name <- name_standardize(name);  standard_name

```

To further format the original `name` vector to the format of first_name last_name, apply the `remove_middle_name` function to the standardized names now in the `standard_name` vector.

```{r, eval = TRUE}

formatted_name <- remove_middle_name(standard_name);  formatted_name

```


####(b) Construct a logical vector indicating whether a character has a title.

Let's standardize the name vector but leave title in place:

```{r, eval = TRUE}
name_with_title <- name_standardize(name, FALSE);  name_with_title

```

Find any character names that include a title:

```{r, eval = TRUE}
# look for the occurence of a string of 2 - 3 characters followed by a period

title <- str_detect(name, '^[:alpha:]{2,3}\\.'); title

```
We should see TRUE for Rev. Timothy Lovejoy and Dr. Julius Hibbert

####(c) Construct a logical vector indicating whether a character has a second name.

This regular expression starts with a 1) positive lookbehind assertion to match the the first name ending with either a space or a period and then 2.) match any word between the next 3.) positive lookbehind assertion looking for the next occurence of whitespace.

Let's look at the standardized name vector which does not include title:

```{r, eval = TRUE}

standard_name

```

Find any name that includes a middle name:

```{r, eval = TRUE}
# use the standard_name vector in the example

middle_name <- str_detect(standard_name, "(?<=. ).*?(?=\\s)"); middle_name

```
The logical vector returned should be TRUE for the name C. Montgomery Burns


###7. Consider the string `<title>+++BREAKING NEWSS+++</title>`.  We would like to extract the first HTML tag.  To do so, we write the regular expression <.+>.  Explain why this fails and correct the expression.

```{r, eval=T}
tag <- '<title>+++BREAKING NEWSS+++</title>'

str_extract(tag, "<.+>")
            
```

This regular expression extracts the entire string because:

`.` -- matches any character except a new line  
`+` -- repeat the previous step from 1 to infinite times, which returns the entire string  


To adjust the regular expression to extract only `<title>`, include `?` so that the shorted match is made once.

```{r, eval=T}

str_extract(tag, "<.+?>")
            
```            

###8. Consider the string `(5-3)^2=5^2-2*5*3+3^2 conforms to the binomial theorem`.  We would like to extract the formula in the string.  To do so, we write the regular expression [^0-9=+*()]+.  Explain why this fails and correct the expression.

```{r, eval = TRUE}

string <- "(5-3)^2=5^2-2*5*3+3^2 conforms to the binomial theorem"

str_extract(string, "[^0-9=+*()]+")

```

This fails because adding the caret `^` inverted the character class when in fact the regular expression needs to include the characters like ^, =, *, +, and ()

changing the regular expression to this will return the formular portion of the string:

```{r, eval = TRUE}
str_extract(string, "[[:digit:]=+*()^-]+")

```

Moving the `^` to the end of the expression along with the other special characters to include in the match changes the expression so that extracts the formula.
